Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-12-02 Tutorial_2a.Rmd
update ms. and tuts

Perform a prior predictive check to see if the binary observations generated from the specified prior distributions reflects our prior beliefs. First, sample the prior distributions using sample_prior="only".

@@

model_M0i_prior <-                  
   brm(data = data_M0i,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + timebin + (0 + timebin | pid),
       prior = priors_M0i,
       chains = 4, cores = 4, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0i_prior")
```

@@

Next, use them to predict prior data. To better understand what the function add_predicted_draws is doing (which we will typically use), we first simulate prior data for one timebin manually.

@@

# Specify
N_samples <- 1000 # the number of samples from the skew_normal
N_obs <- 100      # the number of simulated observations per sample
set.seed(1)

@@

# apply the function
prior_pred <- cloglog_predictive_distribution(cloglog_samples,N_obs)

@@

# plot 
ggplot(prior_pred_haz, aes(x=pred_haz)) +
  geom_histogram(binwidth=.01) +
  scale_x_continuous(limits = c(0,1)) +
  labs(x = "simulated hazard",
       title = "Prior predictive distribution",
       subtitle = str_c(N_samples," samples; ",N_obs," observations per sample"))
```

@@

The shape of the prior predictive distribution looks as expected (compare with Figure 16 (row F) in the supplementary material): hazard values below .5 are more likely than hazard values above .5.

Now use add_predicted_draws() to simulate prior-based data for 10 bins and 6 subjects, while take into account samples from the priors for the standard deviation of the random effects, and correlations between parameters.

@@

# plot 40 draws per participant
ggplot(data = df_pred_haz %>% filter(.draw < 41), 
       aes(x=timebin, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   geom_line(data=df_pred_haz %>% filter(.draw == 20), 
             color="red", 
             linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(6,15), breaks = c(6:15)) +
  labs(x = "time bin",
       y = "simulated hazard",
       title = "Prior predictive distributions",
       subtitle = str_c("8000 samples; ",N_obs," observations per sample")) +
  facet_wrap(~pid)
```

@@

Now we can fit model M0i to get the posterior distributions.

@@

```{r inpect-M0d}
model_M0d <- readRDS("Tutorial_2_Bayesian/models/model_M0d.rds")
formula(model_M0d)
fixef(model_M0d)
```

@@

priors_M0c <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0.1,.1)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.06)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, 1)", class = "sd"),  
  set_prior("lkj(2)", class = "cor")  
)
```

Perform a prior predictive check to see if the binary observations generated from the specified prior distributions reflect our prior beliefs. 
The following code allows to change the coefficients and generate (cloglog-)hazard functions.
Once you have found the coefficients for the intercept, period_9 and period_9_sq that generate a variety of hazard functions spanning your prior beliefs, you can then use them to determine a suited mean and standard deviation for each coefficient.

```{r search-coef}
dat <- tibble(x = -3:6,
              x2 = x^2,
              cloglog1 = -1.8   + 0.3*x   - 0.1*x2,
              cloglog2 = -1     + 0.4*x   - 0.1*x2,
              cloglog3 = -3     - 0.1*x   + 0.12*x2,
              cloglog4 = -0.40  + 0.2*x   - 0.04*x2,
              haz1 = 1 - exp(-1*exp(cloglog1)),
              haz2 = 1 - exp(-1*exp(cloglog2)),
              haz3 = 1 - exp(-1*exp(cloglog3)),
              haz4 = 1 - exp(-1*exp(cloglog4)))

@@

Next, sample the prior distributions, and set the variance of the normal prior for class "sd" to a low value, to minimize the effect of the random effects while inspecting the relation between the selected priors for period_9 and period_9_sq and the shape of the prior predictive distributions.

```{r priors-M0c-ppc}
priors_M0c_ppc <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0.1,.1)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.06)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, .1)", class = "sd"),  # set standard deviation to a small value for now
  set_prior("lkj(2)", class = "cor")  
)
```

```{r fit-model-M0c-prior, eval=F}
plan(multicore)

model_M0c_prior <-
   brm(data = data_M0c,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + Intercept + period_9 + period_9_sq +  
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_ppc,
       chains = 4, cores = 4, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0c_prior")
```

```{r load-M0c-prior}
model_M0c_prior <- readRDS("Tutorial_2_Bayesian/models/model_M0c_prior.rds")
```

@@

```{r prior-distr}
# extract prior draws
post <- as_draws_df(model_M0c_prior) %>% 
  select(-lp__) %>% 
  as_tibble()

@@

```{r ppc-model-M0c-prior, eval=F}
# Generate prior predictive hazard functions for 6 participants
# Specify
N_obs = 100 # increase for smoother results
set.seed(1)

# make new data
newdata_prior = tibble(period_9 = c(-3:6)) %>% 
  mutate(period_9_sq = period_9^2) %>%
  expand_grid(pid = 1:6)

@@

# extract predictions
for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0c_prior, 
                                    newdata = newdata_prior, 
                                    summary=F, 
                                    ndraws=NULL) %>% 
                mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}

@@

# plot 200 draws per participant, highlighting three
sel = sample(1:200,3,replace=F)

ggplot(df_pred_haz %>% 
   filter(.draw <200), aes(x=period_9, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   # highlight
   geom_line(data=df_pred_haz %>% filter(.draw ==sel[1]), 
             color="red", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[2]), 
            color="green", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[3]), 
          color="blue", linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(-3,6), breaks = c(-3:6)) +
  facet_wrap(~pid)
```

```{r save-ppc-M1c, eval=F}
ggsave("Tutorial_2_Bayesian/figures/M0c_ppc.png", width = 10, height = 8, dpi = 800)
```

@@

Now increase the standard deviation again of the normal prior for class "sd", and fit model M0c to get the posterior distributions.

@@

Model_M0c took about 71 minutes to run.

@@

priors_M1d <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"), 
  set_prior("normal(0.1,.1)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.06)", class= "b", coef = "period_9_sq"),  
  set_prior("normal(0,.4)", class = "b", coef = "conditioncongruent"),
  set_prior("normal(0,.4)", class = "b", coef = "conditionincongruent"),
  set_prior("normal(0, 1)", class = "sd"),            
  set_prior("lkj(2)", class = "cor")                  
)
```

@@

Model_M1d took about 145 minutes to run. Note that duplicate terms in the model formula (e.g., condition) are ignored.

```{r inspect-M1d}
model_M1d <- readRDS("Tutorial_2_Bayesian/models/model_M1d.rds")
summary(model_M1d)
formula(model_M1d)
```

@@

# set priors
priors <- c(
  prior(skew_normal(-1,1,-2), class = b, nlpar = a), 
  prior(normal(0.1,.2), class = b, nlpar = b), 
  prior(normal(0,.7), class = b, nlpar = c), 
  prior(exponential(1), class = sd, group = pid, nlpar = a),
  prior(exponential(1), class = sd, group = pid, nlpar = b),
  prior(exponential(1), class = sd, group = pid, nlpar = c),
  prior(lkj(2), class = cor, group = pid)
)
```

@@

@@

We can plot the expected value of the posterior predictive distribution -- the predicted hazard values -- for the average participant, for each participant in the data set, and for a brand new hypothetical participant.

@@

We are actually interested in the difference in predicted hazard between congruent and blank primes on the one hand, and between incongruent and blank primes on the other hand, for each time bin. When based on the grand mean, this is known as the grand average marginal effect (AME).

>>>>

### Intermezzo: Prior Predictive Check for M0i.

Following Gelman et al. (2020), we perform a prior predictive check to see if the binary observations generated from the specified prior distribution(s) reflects our prior beliefs. First, sample the prior distributions using sample_prior="only".

@@

model_M0i_prior <-                  
   brm(data = data_M0i,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + timebin + (0 + timebin | pid),
       prior = priors_M0i,
       chains = 1, cores = 1, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0i_prior")
```

@@

Next, use them to predict prior data. But to better understand what the function add_predicted_draws is doing (which we will typically use), we first simulate prior data for one timebin manually.

@@

# Specify
N_samples <- 2000 # the number of samples from the skew_normal
N_obs <- 100      # the number of simulated observations per sample
set.seed(1)

@@

# apply the function
prior_pred <- cloglog_predictive_distribution(cloglog_samples,
                                              N_obs)

@@

# plot 
ggplot(prior_pred_haz, aes(x=pred_haz)) +
  geom_histogram(binwidth=.01) +
  scale_x_continuous(limits = c(0,1)) +
  labs(x = "simulated hazard",
       title = "Prior predictive distribution",
       subtitle = str_c(N_samples," samples; ",
                        N_obs," observations per sample"))
```

@@

```{r plot-ppc-1-bin, out.width='80%'}
knitr::include_graphics("Tutorial_2_Bayesian/figures/M0i_ppc_1bin.png")
```

The shape of the prior predictive distribution for 1 bin looks as expected (compare with Figure 16 (row F) in the supplementary material): hazard values below .5 are more likely than hazard values above .5.

Now use add_predicted_draws() to simulate prior-based data for 10 bins and 6 subjects, while taking into account samples from the priors for the standard deviation of the random effects, and correlations between parameters.

@@

# plot 40 draws per participant
ggplot(data = df_pred_haz %>% filter(.draw < 200), 
       aes(x=timebin, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   geom_line(data=df_pred_haz %>% filter(.draw == 20), 
             color="red", 
             linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(6,15), breaks = c(6:15)) +
  labs(x = "time bin",
       y = "simulated hazard",
       title = "Prior predictive distributions",
       subtitle = str_c("8000 samples; ",N_obs," observations per sample")) +
  facet_wrap(~pid)
```

@@

```{r plot-ppc-M0i, out.width='80%'}
knitr::include_graphics("Tutorial_2_Bayesian/figures/M0i_ppc.png")
```

The red hazard function is one of the 200 functions plotted, for each of six participants. They can take on any shape, with values below .5 being more likely than values above .5.

After our prior predictive check, we can fit model M0i to get the posterior distributions.

@@

```{r inpect-M0d}
model_M0d <- readRDS("Tutorial_2_Bayesian/models/model_M0d.rds")
fixef(model_M0d)
```

@@

priors_M0c <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0.3,.25)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.06)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, 1)", class = "sd"),  
  set_prior("lkj(2)", class = "cor")  
)
```

### Intermezzo. Prior predictive check for model M0c.

Perform a prior predictive check to see if the binary observations generated from the specified prior distributions reflect our prior beliefs.

To get an idea of how the intercept and both slopes control the shape of the hazard function, the following code allows to change the coefficients and generate (cloglog-)hazard functions.
Once you have found the coefficients for the intercept, period_9 and period_9_sq that generate a variety of hazard functions spanning your prior beliefs, you can then use them to determine a suited mean and standard deviation for each coefficient.

```{r search-coef}
dat <- tibble(x = -3:6,
              x2 = x^2,
              cloglog1 = -1.8   + 0.3*x    - 0.1*x2,  # black
              cloglog2 = -1     + 0.41*x   - 0.04*x2,  # green
              cloglog3 = -3     - 0.13*x   + 0.12*x2, # red
              cloglog4 = -0.40  + 0.6*x    - 0.11*x2, # blue
              haz1 = 1 - exp(-1*exp(cloglog1)),
              haz2 = 1 - exp(-1*exp(cloglog2)),
              haz3 = 1 - exp(-1*exp(cloglog3)),
              haz4 = 1 - exp(-1*exp(cloglog4)))

@@

For example, the coefficients for x vary from -0.13 to 0.6, and we might specify N(.3,.25) as a prior distribution, Those for x-squared vary from -0.11 to 0.12, and we might specify N(0,.06) as a prior distribution.

Suppose we expect (cloglog-)hazard functions that increase almost linearly with time. We also set the standard deviation of the normal prior for class "sd" to a low value, to minimize the effect of the random effects while inspecting the relation between the selected priors for period_9 and period_9_sq and the shape of the prior predictive distributions.

```{r priors-M0c-ppc1}
priors_M0c_ppc1 <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0.3,.001)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.001)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, .01)", class = "sd"), 
  set_prior("lkj(2)", class = "cor")  
)
```

Sample the prior distributions.

```{r fit-model-M0c-prior1, eval=F}
plan(multicore)

model_M0c_prior1 <-
   brm(data = data_M0c,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + Intercept + period_9 + period_9_sq +  
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_ppc1,
       chains = 1, cores = 1, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0c_prior1")
```

```{r load-M0c-prior1}
model_M0c_prior1 <- readRDS("Tutorial_2_Bayesian/models/model_M0c_prior1.rds")
```

@@

```{r prior1-distr}
# extract prior draws
post <- as_draws_df(model_M0c_prior1) %>% 
  select(-lp__) %>% 
  as_tibble()

@@

Create a function.

```{r function-plot-ppc, eval=F}
plot_ppc <- function(fit_prior, N_samples, N_obs, new_data_prior){

set.seed(1)

@@

# extract predictions
for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(fit_prior, 
                                    newdata = new_data_prior, 
                                    summary=F, 
                                    ndraws=N_samples) %>% 
                mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}

@@

# plot N_samples draws per participant, highlighting three
sel = sample(1:N_samples,3,replace=F)

ggplot(df_pred_haz %>% 
   filter(.draw <N_samples), aes(x=period_9, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   # highlight
   geom_line(data=df_pred_haz %>% filter(.draw ==sel[1]), 
             color="red", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[2]), 
            color="green", linewidth=2) +
   geom_line(data=df_pred_haz %>% filter(.draw == sel[3]), 
          color="blue", linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(-3,6), breaks = c(-3:6)) +
  facet_wrap(~pid)
}
```

Apply the function.

```{r eval=F}
# make new data
newdata_prior = tibble(period_9 = c(-3:6)) %>% 
  mutate(period_9_sq = period_9^2) %>%
  expand_grid(pid = 1:6)

# make ppc plot
p1 <- plot_ppc(model_M0c_prior1,2000,100,newdata_prior)
p1
```

For each of 2000 draws from the prior distributions, a black hazard function is plotted. Three randomly selected hazard functions are highlighted in red, green, and blue.

```{r save-ppc1-M0c, eval=F}
ggsave("Tutorial_2_Bayesian/figures/M0c_ppc1.png", width = 10, height = 8, dpi = 800)
```

```{r plot-ppc1-M0c, out.width='80%'}
knitr::include_graphics("Tutorial_2_Bayesian/figures/M0c_ppc1.png")
```

Or suppose we believe apriori that the hazard functions are flat.
We specify our new priors:

```{r priors-M0c-ppc2}
priors_M0c_ppc2 <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0,.001)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.001)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, .001)", class = "sd"), 
  set_prior("lkj(2)", class = "cor")  
)
```

Sample the prior distributions 2000 times:

```{r fit-model-M0c-prior2, eval=F}
plan(multicore)

model_M0c_prior2 <-
   brm(data = data_M0c,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + Intercept + period_9 + period_9_sq +  
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_ppc2,
       chains = 1, cores = 1, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0c_prior2")
```

```{r load-M0c-prior2}
model_M0c_prior2 <- readRDS("Tutorial_2_Bayesian/models/model_M0c_prior2.rds")
```

Apply the function plot_ppc().

```{r eval=F}
# make new data
newdata_prior = tibble(period_9 = c(-3:6)) %>% 
  mutate(period_9_sq = period_9^2) %>%
  expand_grid(pid = 1:6)

# make ppc plot
p2 <- plot_ppc(model_M0c_prior2,2000,100,newdata_prior)
p2
```

Now the prior predictive hazard functions are all rather flat, with an intercept controlled by the prior skew-normal distribution, i.e., values below .5 are more likely than values above .5.

```{r save-ppc2-M0c, eval=F}
ggsave("Tutorial_2_Bayesian/figures/M0c_ppc2.png", width = 10, height = 8, dpi = 800)
```

```{r plot-ppc2-M0c, out.width='80%'}
knitr::include_graphics("Tutorial_2_Bayesian/figures/M0c_ppc2.png")
```

Finally, perform a ppc for the selected prior distributions for M0c (see lines 369-374), and increase the standard deviation for the normal prior for class "sd" somewhat.

```{r priors-M0c-ppc3}
priors_M0c_ppc3 <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0.3,.25)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.06)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, .1)", class = "sd"),  
  set_prior("lkj(2)", class = "cor")  
)
```

```{r fit-model-M0c-prior3, eval=F}
plan(multicore)

model_M0c_prior3 <-
   brm(data = data_M0c,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + Intercept + period_9 + period_9_sq +  
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_ppc3,
       chains = 1, cores = 1, 
       iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, 
                      step_size = 0.04, 
                      max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0c_prior3")
```

```{r load-M0c-prior3}
model_M0c_prior3 <- readRDS("Tutorial_2_Bayesian/models/model_M0c_prior3.rds")
```

Apply the function plot_ppc().

```{r eval=F}
# make new data
newdata_prior = tibble(period_9 = c(-3:6)) %>% 
  mutate(period_9_sq = period_9^2) %>%
  expand_grid(pid = 1:6)

# make ppc plot
p3 <- plot_ppc(model_M0c_prior3,2000,100,newdata_prior)
p3
```

@@

```{r save-ppc3-M0c, eval=F}
ggsave("Tutorial_2_Bayesian/figures/M0c_ppc3.png", width = 10, height = 8, dpi = 800)
```

```{r plot-ppc3-M0c, out.width='80%'}
knitr::include_graphics("Tutorial_2_Bayesian/figures/M0c_ppc3.png")
```

After these prior predictive checks, we can increase the standard deviation for class "sd" to 1 to cover many possible inter-individual differences, and fit model M0c to get the posterior distributions.

@@

Model_M0c took about 19 minutes to run.

@@

priors_M1d <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"), 
  set_prior("normal(0.3,.25)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.06)", class= "b", coef = "period_9_sq"),  
  set_prior("normal(0,.4)", class = "b", coef = "conditioncongruent"),
  set_prior("normal(0,.4)", class = "b", coef = "conditionincongruent"),
  set_prior("normal(0, 1)", class = "sd"),            
  set_prior("lkj(2)", class = "cor")                  
)
```

@@

Model_M1d took about 184 minutes to run. Note that duplicate terms in the model formula (e.g., condition) are ignored.

```{r inspect-M1d}
model_M1d <- readRDS("Tutorial_2_Bayesian/models/model_M1d.rds")
summary(model_M1d)
```

@@

# set priors
priors <- c(
  prior(skew_normal(-1,1,-2), class = b, nlpar = a), 
  prior(normal(0.3,.25), class = b, nlpar = b), 
  prior(normal(0,.06), class = b, nlpar = c), 
  prior(exponential(1), class = sd, group = pid, nlpar = a),
  prior(exponential(1), class = sd, group = pid, nlpar = b),
  prior(exponential(1), class = sd, group = pid, nlpar = c),
  prior(lkj(2), class = cor, group = pid)
)
```

@@

Observations with Pareto k values above .7 are problematic.

@@

Following Heiss (2021), we can plot the expected value of the posterior predictive distribution -- the predicted hazard values -- for the average participant, for each participant in the data set, and for a brand new hypothetical participant.

@@

Following Heiss (2021), we are actually interested in the difference in predicted hazard between congruent and blank primes on the one hand, and between incongruent and blank primes on the other hand, for each time bin. When based on the grand mean, this is known as the grand average marginal effect (AME).

### 2024-12-02 manuscript/SupplementaryMaterial.Rmd
update ms. and tuts

We defined 13 custom functions that we list here.

* censor(df,timeout,bin_width) : divide the time segment (0,timeout] in bins, identify any right-censored observations, and determine the discrete RT (time bin rank)
* ptb(df) : transform the person-trial data set to the person-trial-bin data set
* setup_lt(ptb) : set up a life table for each level of 1 independent variable
* setup_lt_2IV(ptb) : set up a life table for each combination of levels of 2 independent variables
* calc_ca(df) : estimate the conditinal accuracies when there is 1 independent variable
* calc_ca_2IV(df) : estimate the conditional accuraies when there are 2 independent variables
* join_lt_ca(df1,df2) : add the ca(t) estimates to the life tables (1 independent variable)
* join_lt_ca_2IV(df1,df2) : add the ca(t) estimates to the life tables (2 independent variables)
* extract_median(df) : estimate quantiles S(t)~.50~ (1 independent variable)
* extract_median_2IV(df) : estimate quantiles S(t)~.50~ (2 independent variables)
* plot_eha(df,subj,haz_yaxis=1,first_bin_shown=1,aggregated_data=F,Nsubj=6) : create plots of the discrete-time functions (1 independent variable), and specify the upper limit of the y-axis in the hazard plot, with which bin to start plotting, whether the data is aggregated across participants, and across how many participants
* plot_eha_2IV(df,subj,haz_yaxis=1,first_bin_shown=1,aggregated_data=F,   Nsubj=6) : create plots of the discrete-time functions (2 independent variables), and specify the upper limit of the y-axis in the hazard plot, with which bin to start plotting, whether the data is aggregated across participants, and across how many participants

@@

An example (single-level) discrete-time hazard model with three predictors (TIME, X~1~, X~2~), the cloglog link function, and a third-order polynomial specification for TIME can be written as follows:

@@

The main predictor variable TIME is the time bin index t that is centered on value 9 in this example. The first set of terms within brackets, the parameters $\beta$~0~ to $\beta$~2~ multiplied by their polynomial specifications of (centered) time, represents the shape of the baseline cloglog-hazard function (i.e., when all predictors X~i~ take on a value of zero). The second set of terms (the beta parameters $\beta$~3~ to $\beta$~5~) represents the vertical shift in the baseline cloglog-hazard for a 1 unit increase in the respective predictor variable. Predictors can be discrete, continuous, and time-varying or time-invariant. For example, the effect of a 1 unit increase in X~1~ is to vertically shift the whole baseline cloglog-hazard function by $\beta$~1~ cloglog-hazard units. However, if the predictor interacts linearly with TIME (see X~2~ in the example), then the effect of a 1 unit increase in X~2~ is to vertically shift the predicted cloglog-hazard in bin 9 by $\beta$~2~ cloglog-hazard units (when TIME-9 = 0), in bin 10 by $\beta$~2~ + $\beta$~3~ cloglog-hazard units (when TIME-9 = 1), and so forth. To interpret the effects of a predictor, its $\beta$ parameter is exponentiated, resulting in a hazard ratio (due to the use of the cloglog link). When using the logit link, exponentiating a $\beta$ parameter results in an odds ratio.

@@

To gain a sense of what prior *logit* values would approximate a uniform distribution on the probability scale, @kurzAppliedLongitudinalDataAnalysis2023 simulated a large number of draws from the Uniform(0,1) distribution, converted those draws to the log-odds metric, and fitted a Student's t distribution. Row C in Figure 12 shows that using a t-distribution with 7.61 degrees of freedom and a scale parameter of 1.57 as a prior on the logit scale, approximates a uniform distribution on the probability scale. According to @kurzAppliedLongitudinalDataAnalysis2023, such a prior might be a good prior for the intercept(s) in a logit-hazard model, while the N(0,1) prior in row D might be a good prior for the non-intercept parameters in a logit-hazard model, as it gently regularizes p towards .5 (i.e., a zero effect on the logit scale).

@@

To gain a sense of what prior *cloglog* values would approximate a uniform distribution on the hazard probability scale, we followed Kurz's approach and simulated a large number of draws from the Uniform(0,1) distribution, converted them to the cloglog metric, and fitted a skew-normal model (due to the asymmetry of the cloglog link function). Row E shows that using a skew-normal distribution with a mean of -0.59, a standard deviation of 1.26, and a skewness of -4.22 as a prior on the cloglog scale, approximates a uniform distribution on the probability scale. 
However, because hazard values below .5 are more likely in RT studies, using a skew-normal distribution with a mean of -1, a standard deviation of 1, and a skewness of -2 as a prior on the cloglog scale (row F), might be a good weakly informative prior for the intercept(s) in a cloglog-hazard model. 
A skew-normal distribution with a mean of -0.2, a standard deviation of 0.71, and a skewness of -2.2 might be a good weakly informative prior for the non-intercept parameters in a cloglog-hazard model as it gently regularizes p towards .6321 (i.e., a zero effect on the cloglog scale).

@@

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by @holdenDispersionResponseTimes2009, “probability density [and mass] functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions” (p. 331) when one is interested in studying the detailed shape of a RT distribution [see also Figure 1 in @panisAnalyzingResponseTimes2020]. Therefore, when the goal is to study how psychological effects change over time, hazard and conditional accuracy functions are preferred.

>>>>

We defined 12 custom functions that we list here.

* censor(df,timeout,bin_width) : divide the time segment (0,timeout] in bins, identify any right-censored observations, and determine the discrete RT (time bin rank)
* ptb(df) : transform the person-trial data set to the person-trial-bin data set
* setup_lt(ptb) : set up a life table for each level of 1 independent variable
* setup_lt_2IV(ptb) : set up a life table for each combination of levels of 2 independent variables
* calc_ca(df) : estimate the conditinal accuracies when there is 1 independent variable
* calc_ca_2IV(df) : estimate the conditional accuraies when there are 2 independent variables
* join_lt_ca(df1,df2) : add the ca(t) estimates to the life tables (1 independent variable)
* join_lt_ca_2IV(df1, df2) : add the ca(t) estimates to the life tables (2 independent variables)
* extract_median(df) : estimate quantiles S(t)~.50~ (1 independent variable)
* extract_median_2IV(df) : estimate quantiles S(t)~.50~ (2 independent variables)
* plot_eha(df, subj, haz_yaxis=1, first_bin_shown=1, aggregated_data=F, Nsubj=6) : create plots of the discrete-time functions (1 independent variable), and specify the upper limit of the y-axis in the hazard plot, with which bin to start plotting, whether the data is aggregated across participants, and across how many participants
* plot_eha_2IV(df, subj, haz_yaxis=1, first_bin_shown=1, aggregated_data=F,   Nsubj=6) : create plots of the discrete-time functions (2 independent variables), and specify the upper limit of the y-axis in the hazard plot, with which bin to start plotting, whether the data is aggregated across participants, and across how many participants

@@

An example (single-level) discrete-time hazard model with three predictors (TIME, X~1~, X~2~), the cloglog link function, and a second-order polynomial specification for TIME can be written as follows:

@@

The main predictor variable TIME is the time bin index t that is centered on value 9 in this example. The first set of terms within brackets, the parameters $\beta$~0~ to $\beta$~2~ multiplied by their polynomial specifications of (centered) time, represents the shape of the baseline cloglog-hazard function (i.e., when all predictors X~i~ take on a value of zero). The second set of terms (the beta parameters $\beta$~3~ to $\beta$~5~) represents the vertical shift in the baseline cloglog-hazard for a 1 unit increase in the respective predictor variable. Predictors can be discrete, continuous, and time-varying or time-invariant. For example, the effect of a 1 unit increase in X~1~ is to vertically shift the whole baseline cloglog-hazard function by $\beta$~3~ cloglog-hazard units. However, if the predictor interacts linearly with TIME (see X~2~ in the example), then the effect of a 1 unit increase in X~2~ is to vertically shift the predicted cloglog-hazard in bin 9 by $\beta$~4~ cloglog-hazard units (when TIME-9 = 0), in bin 10 by $\beta$~4~ + $\beta$~5~ cloglog-hazard units (when TIME-9 = 1), and so forth. To interpret the effects of a predictor, its $\beta$ parameter is exponentiated, resulting in a hazard ratio (due to the use of the cloglog link). When using the logit link, exponentiating a $\beta$ parameter results in an odds ratio.

@@

To gain a sense of what prior *logit* values would approximate a uniform distribution on the probability scale, @kurzAppliedLongitudinalDataAnalysis2023 simulated a large number of draws from the Uniform(0,1) distribution, converted those draws to the log-odds metric, and fitted a Student's t distribution. Row C in Figure 16 shows that using a t-distribution with 7.61 degrees of freedom and a scale parameter of 1.57 as a prior on the logit scale, approximates a uniform distribution on the probability scale. According to @kurzAppliedLongitudinalDataAnalysis2023, such a prior might be a good prior for the intercept(s) in a logit-hazard model, while the N(0,1) prior in row D might be a good prior for the non-intercept parameters in a logit-hazard model, as it gently regularizes p towards .5 (i.e., a zero effect on the logit scale).

@@

To gain a sense of what prior *cloglog* values would approximate a uniform distribution on the hazard probability scale, we followed Kurz's approach and simulated a large number of draws from the Uniform(0,1) distribution, converted them to the cloglog metric, and fitted a skew-normal model (due to the asymmetry of the cloglog link function). Row E shows that using a skew-normal distribution with a mean of -0.59, a standard deviation of 1.26, and a skewness of -4.22 as a prior on the cloglog scale, approximates a uniform distribution on the probability scale. 
However, because hazard values below .5 are more likely in RT studies, using a skew-normal distribution with a mean of -1, a standard deviation of 1, and a skewness of -2 as a prior on the cloglog scale (row F), might be a good weakly informative prior for the intercept(s) in a cloglog-hazard model.

@@

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by @holdenDispersionResponseTimes2009, “probability density [and mass] functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions” (p. 331) when one is interested in studying the detailed shape of a RT distribution [see also Figure 1 in @panisAnalyzingResponseTimes2020]. Therefore, when the goal is to study how psychological effects change over time, hazard and conditional accuracy functions are the preferred ways to describe the RT + accuracy data.
